Remote sensing riverine gas exchange and estimating carbon efflux using SWOT observations

### 2021-02-19 manuscript_v1.Rmd
manuscript edits

$$Flux=k(C_{water}-C_{air}) \text{(1)}$$

It is infeasible to measure $k_{600}$ directly when working at the drainage-network-scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007]. This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976]. Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by estimates of the *V* of the river channel. Thus, drainage-network-scale GHG emissions estimates are highly sensitive to their *V* data. This is exasperated in ungauged basins where *V* models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

@@

Remote hydraulic estimation is itself a burgeoning subfield within remote sensing of hydrology and an analogue to the problem presented here is the remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review). There are many ways to perform RSQ, from calibrating to channel hydraulics [e.g. @brakenridgeOrbitalMicrowaveMeasurement2007; @pavelskyUsingWidthbasedRating2014; @pavelskyRemoteSensingSuspended2009; @tarpanelliEstimationRiverDischarge2013] to calibrating hydraulic/hydrologic models [e.g. @bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information as inputs to  hydraulic/hydrologic models (termed 'ungauged approaches'). Many of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.

@@

Thus, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales, after techniques have been devleoped to ingest its novel measurements of water surface width and height. Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information. We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory. More specifically, this manuscript aims to answer two questions:
	1) Is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements?
	2) How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?

@@

##2 Methods
We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’ algorithm. To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that relies on no a priori knowledge of the river (section 2.1) and then we couple that upscaling model with remote sensing techniques to build BIKER (section 2.2). Following these methods, the validation setup is described (section 2.3), as is the comparison of estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4). A flowchart detailing the complete workflow is included (Figure 1).

@@

###2.1 Developing an RS-able $k_{600}$ upscaling model
[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient in a similar manner [e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976]. Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low. Their upscaling function is reprinted as equation 2.

$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  \text{(2)}$$
$$ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$$

While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters. We do not have this a priori knowledge in an ungauged, RS-based setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width to define river size and train different upscaling functions for different groups of data with similar river sizes [Figure 2, using the data from @ulsethDistinctAirWater2019a]. In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

$$k_{600}=a_i(eD)^{b_i} \text{(3)}$$

@@

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$. The approach used here is strongly informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize remote sensing $k_{600}$ as a Bayesian inference problem.

@@

In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function. Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ for the change in channel area *dA* and median channel area $A_0$). It is also defined by the upscaling parameters $a_i$ and $b_i$.

$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} \text{(4)}$$

Equation 4 is then reworked into a Bayesian sampling model with the ‘data’ on the left-hand side sampled from the unknown model parameters on the right-hand side (equation 5). Note that in the formal likelihood specification each term in equation 5 is log-transformed to yield a normal sampling model, however they are written here without a transformation for succinctness. $σ_{upscale}^2$ refers to the uncertainty inherent in the $k_{600}$ estimates themselves (equation 4).

@@

Equation five necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$. Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters. Or more intuitively, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < x < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter $X=n$, $A_0$, or $k_{600}$. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. *n* and $A_0$ prior hyperparameters were assigned following the method developed by [@brinkerhoffConstrainingRemoteRiver2020]. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the [@ulsethDistinctAirWater2019a] dataset to assign $\mu$ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   \text{(6)}$$
$$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$$

Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. Specifically, We explore three uncertainty scenarios to assess which source is most influential. Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error. Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes. To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics under three different uncertainty scenarios [@brinkerhoffReconcilingAtaStationAtManyStations2019]. Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3). Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25. The specifics of the slope-error model used are elaborated in Section 2.3 [@durandHowWillRadar2020a]. This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term is extracted and used for $σ_{upscale}^2$.

@@

###2.3 BIKER validation setup
We validated the BIKER algorithm on 22 rivers using observed $V$ and observed $S_h$ First, we detail the 22 rivers, and then we detail how ‘observed’ $k_{600}$ is calculated.

@@

We also introduced measurement error into these simulated SWOT observations to assess BIKER's performance degradation due to errors inherent in SWOT measurements. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error. Here, we assume errors in river width are negligible [@durandHowWillRadar2020a] and solely focus on measurement errors in river height and slope. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, SOMETHING HERE. This generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013] where the land surface 'obscures' the water surface relative to SWOT's incidence angle. While the SWOT simulator mentioned above is ideal for this analysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT radar layover errors [@durandHowWillRadar2020a]. They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches. The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively. To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from normal distributions following equations 7 and 8. Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a] and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$.

$$S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2}) (7)$$

$$H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2}) (8)$$

@@

$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \text{(9)}$$

@@

First, we calculated $FCO_2$. To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations and water temperatures. 26 bi-weekly samples were made by @beaulieuControlsGasTransfer2012a- Figure S1) at one location in the Ohio River for one calender year from 2008-2009. We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and this was deemed acceptable). Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st. Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. Atmospheric $CO_2$ was assumed 390 uatm. The Schmidt number, used to obtain $k_{CO2}$ from $k_{600}$, was calculated following [@raymondScalingGasTransfer2012a] and [@wanninkhofRelationshipWindSpeed1992a].

Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three gauge-based HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a]. See Table S1 for their definitions. It is worth stressing that the [@lauerwaldSpatialPatternsCO22015] model was developed as one of two components of the [@raymondGlobalCarbonDioxide2013a] model, and so was actually developed by the latter’s authors. The names used here refer to the specific implementation used in each study. For our study, all HG methods use the in-situ discharge record while BIKER does not. This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models). Finally, we express the bulk carbon efflux (via $CO_2$ evasion) $[M/T]$ as the average mass flow rate of carbon per year from the 22 rivers, accounting for total river surface area.

##3 Results
First, we present the results from the $k_{600}$ upscaling experiment (section 3.1). We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2). Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates from the 22 rivers across gauged and ungauged methods (section 3.3).

###3.1 Validating the $k_{600}$ upscaling model
Figure 3 plots the validations for the five upscaling functions tested. Both the original [@ulsethDistinctAirWater2019a] model and our RS-able extension perform the best on this independent set of validation data, with near identical performance. The RS-able [@ulsethDistinctAirWater2019a] model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original [@ulsethDistinctAirWater2019a] model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively). The two power-law based [@raymondScalingGasTransfer2012a] models (equations 4 and 3 in that study) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively). [@raymondScalingGasTransfer2012a] equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset. In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed. **SHOULD I INCLUDE THIS MODEL??** Because we successfully reproduced the [@ulsethDistinctAirWater2019a] model, the rest of this manuscript deals exclusively with this RS-able variant of the model.

@@

Uncertainties were then propagated through the RS-able [@ulsethDistinctAirWater2019a] model via MC simulations for three error scenarios (section 2.2) for the 5,000 hydraulic measurements. Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions (Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms). Across the 5,000 tests for the fullest error scenario (Figure 4e), we quantified average ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$medianSigma,2)` (~`r round(exp(results_3_1_MC$medianSigma), 2)` m/day- dashed blue line in Figure 4a). Uncertainty solely from the upscaling model is also 1.27, confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER (specifically, using Manning’s equation to approximate average flow velocity and SWOT measuurement error). This also confirms the large residuals inherent in $k_{600}$ upscaling models as described by @hallGasExchangeStreams2020.

>>>>

$$Flux=k(C_{water}-C_{air}) \mathbf{(1)}$$

It is infeasible to measure $k_{600}$ directly when working at the drainage-network-scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007]. This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976]. Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the *V* of the river channel. This is exasperated in ungauged basins where *V* models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

@@

Remote hydraulic estimation is itself a burgeoning subfield within remote sensing of hydrology and an analogue to the problem presented here is the remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review). There are many ways to perform RSQ, from calibrating RS data to local channel hydraulics [e.g. @brakenridgeOrbitalMicrowaveMeasurement2007; @pavelskyUsingWidthbasedRating2014; @pavelskyRemoteSensingSuspended2009; @tarpanelliEstimationRiverDischarge2013] to calibrating hydraulic/hydrologic models with both in situ and RS data [e.g. @bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information in their hydraulic/hydrologic models (termed 'ungauged approaches'). Many, but not all, of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. **I don't know how if introducing SWOT via RSQ is the best idea. Maybe I should just introduce SWOT on its own?**

@@

In this context, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales after techniques have been developed to ingest SWOT's novel measurements of water surface width and height. Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information. We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory. More specifically, this manuscript aims to answer two questions:
	1) Is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements?
	2) How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?

@@

##2 Methods
We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’ algorithm. To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that relies on no a priori knowledge of river hydraulics (section 2.1) and then we couple that upscaling model with remote sensing techniques to build BIKER (section 2.2). Following these methods, the validation setup is described (section 2.3), as is the workflow for comparing estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4). A flowchart detailing the entire study is included (Figure 1).

@@

###2.1 Developing an RS-able $k_{600}$ upscaling model
[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope. This extended earlier work predicting the reaeration coefficient in a similar manner [e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976]. Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a] by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low. Their upscaling function is reprinted as equation 2.

$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  \mathbf{(2)}$$
$$ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$$

While equation 2 would be ideal for the goal of this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters. We do not have this a priori knowledge in an ungauged, RS-based setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width to define river size and train different upscaling functions for different groups of data with similar river sizes [Figure 2, using the data from @ulsethDistinctAirWater2019a]. In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

$$k_{600}=a_i(eD)^{b_i} \mathbf{(3)}$$

@@

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to remotely sense $k_{600}$. The approach used here is strongly informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize remote sensing $k_{600}$ as a Bayesian inference problem that can be solved using SWOT data.

@@

In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function. Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for ingesting SWOT data). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (for the change in channel area *dA* and median channel area $A_0$). It is also defined by the upscaling parameters $a_i$ and $b_i$.

$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} \mathbf{(4)}$$

Equation 4 is then reworked into a Bayesian sampling model with the ‘data’ on the left-hand side sampled from the unknown model parameters on the right-hand side (equation 5). Note that in the formal likelihood specification each term in equation 5 is log-transformed to yield a normal sampling model, however they are written here without a transformation for succinctness. The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the $k_{600}$ estimates themselves (equation 4).

@@

Equation five necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$. Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters. Or more intuitively for these physcial quantities, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < x < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter $X=n$, $A_0$, or $k_{600}$. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. *n* and $A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the @ulsethDistinctAirWater2019a dataset to assign $\mu$ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   \mathbf{(6)}$$
$$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$$

Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. Specifically, we explore three uncertainty scenarios to assess which source of error is most influential. Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error. Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes. To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019] under three different uncertainty scenarios. Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3). Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25. The specifics of the slope-error model that we used are elaborated on in Section 2.3 and rely on a recent analysis of SWOT slope measurement errors [@durandHowWillRadar2020a]. This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term for each uncertainity scenario is extracted and used for $σ_{upscale}^2$.

@@

###2.3 BIKER validation setup
We validated BIKER on 22 rivers using observed $V$ and observed $S_h$ First, we detail the 22 rivers, and then we detail how ‘observed’ $k_{600}$ is calculated.

@@

We also introduced measurement error into these simulated SWOT observations to assess BIKER's performance degradation due to errors inherent in SWOT measurements. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error. Here, we assume errors in river width are negligible [@durandHowWillRadar2020a] and solely focus on measurement errors in river height and slope. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, SOMETHING HERE. This generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013] where the land surface 'obscures' the water surface relative to SWOT's incidence angle. While the SWOT simulator mentioned above is ideal for this error nalysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a]. They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches. The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively. To apply this model to our 22 rivers, we generate new river heights and water surface slopes by sampling every observation from normal distributions following equations 7 and 8 for node *i* and timestep *t*. Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a] and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$.

$$S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2}) \mathbf{(7)}$$

$$H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2}) \mathbf{(8)}$$

@@

$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \mathbf{(9)}$$

@@

First, we calculated $FCO_2$. To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations and water temperatures. 26 bi-weekly samples were made by @beaulieuControlsGasTransfer2012a- Figure S1) at one location in the Ohio River for one calender year from 2008-2009. We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and this was deemed acceptable). Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st. Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. Atmospheric $CO_2$ was assumed 390 uatm. The Schmidt number, used to obtain $k_{CO2}$ from $k_{600}$, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.

Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three gauge-based HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a]. See Table S1 for their definitions. It is worth stressing that the @lauerwaldSpatialPatternsCO22015 model was developed as one of two components of the @raymondGlobalCarbonDioxide2013a model, and so was actually developed by the latter’s authors. The names used here refer to the specific implementation used in each study. For our study, all HG methods use the in-situ discharge record while BIKER does not. This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models). Finally, we express the bulk carbon efflux (via $CO_2$ evasion) $[M/T]$ as the average mass flow rate of carbon per year from the 22 rivers, accounting for total river surface area.

##3 Results
First, we present the results from the $k_{600}$ upscaling experiment (section 3.1). We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2). Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates made using both gauged and ungauged methods (section 3.3).

###3.1 Validating the $k_{600}$ upscaling model
Figure 3 plots the validations for the five upscaling functions tested. Both the original @ulsethDistinctAirWater2019a model and our RS-able extension perform the best on this independent set of validation data, with near identical performance. The RS-able @ulsethDistinctAirWater2019a model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original @ulsethDistinctAirWater2019a model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively). The two power-law based @raymondScalingGasTransfer2012a models (equations 4 and 3 in that study) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively). @raymondScalingGasTransfer2012a equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset. In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed. **SHOULD I INCLUDE THIS MODEL or not mention it??** Because we successfully reproduced the @ulsethDistinctAirWater2019a model, the rest of this manuscript deals exclusively with this RS-able variant of the model.

@@

Uncertainties were then propagated through the RS-able @ulsethDistinctAirWater2019a model via MC simulations for the three error scenarios (section 2.2) for the 5,000 hydraulic measurements. Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions (Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms). Across the 5,000 tests for the fullest error scenario (Figure 4e), we quantified median ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$medianSigma,2)` (~`r round(exp(results_3_1_MC$medianSigma), 2)` m/day- dashed blue line in Figure 4a). Uncertainty solely from the upscaling model is also 1.27, confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER (specifically, using Manning’s equation to approximate average flow velocity and SWOT measurement error). Figures 4b, 4d, and 4f are also functionally identical, further confirming this result.
